---
title: "CMSC 320- Final Tutorial"
author: "Dipisha Singh"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```
# Get the story behind Olympics Medal 

## Introduction

This tutorial will introduce the entire datascience pipeline from data curation, parsing, and management to exploratory data analysis to hypothesis testing and machine learning.The data taken for this tutorial is from the Olympics Medal list of 2016.The motive to choose this data is to find out the dependencies of medals; for instance, winning medals in olympics can give information about the country's strong and popular sport. It can also give the range of victory in Olympics based on gender, events, record type.

Sports has been medium of international co-operation and the platform where one can represent the country. Olympics is an international multi-sport event and 206 nations particpated in Olympics 2016. There are always economic as well as prestigious benefits for the country on winning the medals. With the data and its analysis, businesses try to get grips for the players with the potetial to win in the next olympic. Therefore, olympics data analysis is necessary for comparison as well as response enactment for better outcome in future.

## Overview and Content
In this tutorial, we will demonstrate
1. Procurement of olympic medals' data via R.
2. Plotting of several variables ad visualization of relationships between medals and other variables
3.Implementatio of hypotheis testing and machine learning through t-statistic and linear regression

## Libraries Used

The libraries used in this project are tidyverse, dplyr, broom, tree, plotrix, ggplot2, stringr.

## Data Curation and Parsing

Data Curation is an important step during the data analysis process. It provides meaningful access to the data.It involves constructing a structure that sorts the data into a meaningful retrieval system and gives us intellectual control over content.

In this tutorial, We will use lot of functions from deifferent libraries as well as we will write function as per the necessity to decrease redundancy of code.
We will create, scrape and parse the given file from the url and make a dataframe for further analyis

### Importing the libraries and reading table from url

```{r reading url}
library(dplyr)
library(tidyr)
library(tidyverse)
library(stringr)
library(ggplot2)
library(plotrix)
library(tree)
library(broom)

# Reading the url
url <- "https://raw.githubusercontent.com/nbremer/olympicfeathers/gh-pages/data/raw%20medal%20data/Rio%202016%20gold%20medal%20winners.txt"
olympics <- read.table(url, sep = "\t", header = TRUE, na.strings = "", 
                       stringsAsFactors = FALSE)

# Giving column names
colnames(olympics) <- c("Athlete_Name", "Country", "Country_Code" , "Gender" , "Sport" , "Discipline" , "Event", "Event_Gender" , "Record" , "Record_Type")

as_data_frame(olympics)
```

### Tidying the data

This section includes data preparation, that is taking raw datasets and make changes to turn messy data into tidy data. If we open the url and see the table, we can observe that there are lots of things that can be made simple. For instance, we can drop the gender column as Gender and Event_gender column has the same thing except Gender has women and men and Event_gender has M ad W as symbols. After dropping we will then rename Event_gender as Gender.On the other hand, it can be observed that Records and Record_Type are mostly NA. As other columns have important information, it is not good to just eliminate athletes with NA record. SO, I have combined Record and record Type as records.

### replace_na_new function

While tidying data, below we will see replace_na_new function which is written to replace any text or number for any columns haing NA.

```{r drop_gender}
# Removing Gender and renaming Event_Gender as Gender
tidy_olympics <- subset(olympics, select= -c(Gender))
colnames(tidy_olympics)[colnames(tidy_olympics)== "Event_Gender"] <- "Gender"

# Writing min as m and sec as s in Record column
tidy_olympics$Record <- gsub('min', 'm', tidy_olympics$Record)
tidy_olympics$Record<- gsub('sec', 's', tidy_olympics$Record)

#Writing Yes if record typeis OR(olympic Record)
tidy_olympics$if_OR <-ifelse(tidy_olympics$Record_Type == "OR", "Yes", "No") 

# Extracting the name of Event and separating 
tidy_olympics$Event_name <- sub("^\\S+\\s+", '', tidy_olympics$Event)
tidy_olympics$Num <- str_extract(tidy_olympics$Event, "[0-9]+")

# replace function which writes the desired string or number for NA 
replace_na_new <- function(data, replace = list(), ...){
  stopifnot(rlang::is_list(replace))
  replace <- rlang::modify(replace, ...)
  for (var in names(replace)) {
    data[[var]][rlang::are_na(data[[var]])] <- replace[[var]]
  }
  data
}

# Using above replace function 
olympics_sub <- tidy_olympics%>%
  subset(select = -c(Event))%>%
  replace_na_new(Athlete_Name = "Anonymous", Record = 0, Record_Type = "None", Num= 0)%>%
  unite(Records, Record, Record_Type, sep='- ')%>%
  arrange(desc(Country))
  

as_data_frame(olympics_sub)
```

## Data Management

Here, we will divide the dataframe into subsets.Getting subsets of olympic_sub which makes it easier for analysis later. We will see the reason in the analysis section. The first one is olympics_data includes specific columns(Country, Sport, Gender, Event_name). Any unnecessary variables are removed. The second dataframe is have _record which includes Country_Code, Sport, Records, if_OR. if_OR gives yes if it is olypic record type and any recordtype, athlete name with NA rows are eliminated in this seocnd dataframe.


```{r olym_dta}
# Dropping Country and renaming country_code with country
olympics_data<- olympics_sub %>%
  subset(select = -c(Athlete_Name, Country, Discipline, Records, if_OR, Num))

colnames(olympics_data)[colnames(olympics_data)== "Country_Code"] <- "Country"
as_data_frame(olympics_data)
```


```{r having_record}
# filtering out the rows with NA
have_record <- olympics_sub%>%
  filter(Records != "0- None" & Athlete_Name != "Anonymous")%>%
  select(Athlete_Name, Gender, Country_Code, Sport, Records, if_OR)

colnames(have_record)[colnames(have_record)== "Country_Code"] <- "Country"
as_data_frame(have_record)
```
  

## Exploratory Data Analysis (EDA)

Exploratory Data Analysis helps to better understand the tidy data, and help us make decisions about appropriate statistical or Machine Learning methods, or data transformations.In most of the cases,statistical data modeling is not required to tell a clear and convincing story;an effective visualization can lead to convincing conclusions.

The goal of EDA in this tutorial is to perform an initial exploration of attributes/variables across entities/observations.We will focus on exploration of single or pair of variables. Length function has been very useful to identify the number of medals based on many variables like gender, country, events. Also, Popular games was investigated for all the participating countries. With the top five countries having highest number of medals, it is necessary to analyze the sport that these countries win most medal for. 

### Number of Medals won By Gender

This plot uses barplot to give number of medals won by men , women and x gender.Itcan be observed that there is not much difference of wins between men and women.

```{r plot_one}
gender_data<- olympics_data%>%
  group_by(Gender)%>%
  summarise(Gender_Count = length(Gender))%>%
  ggplot(aes(x=Gender, y = Gender_Count))+
  geom_bar(stat="identity", mapping = aes(fill= Gender, color=Gender))+
  xlab("Gender") +   
  ylab("Number of Gold medalists") +   
  ggtitle("Number of Gold medalists based on Gender")
gender_data
```
 
### Number of Medals Won by Country and its plot

This plot is based on the country and USA has highest wins.
```{r plot_two, out.width='50%'}
country_count <- olympics_data%>% 
  group_by(Country)%>%
  summarise(wins = length(Country))%>%
  arrange(desc(wins))
country_count
```

This is the scatter plot of above data and it can be seen distributed unsymmetrically.The range of wins is mostly below 10.We will check this in next plot.

```{r plot}
country_count%>%
  ggplot(aes(x=Country, y = wins))+
  geom_point(stat="identity", mapping = aes(fill= Country, color=Country))+
  xlab("Country") +   
  ylab("Number of Gold medalists") +   
  ggtitle("Number of Gold medalists from each country")+
   theme(axis.text.x = element_text(angle=90, hjust=0.5))+
  coord_fixed()
```

The following plot calculates the mean and it is found that the mean is below 10 that is 5.

```{r plot_e}
country_count%>%
  ggplot(aes(x=Country, y = wins))+
  geom_histogram(stat="identity", mapping = aes(fill= Country, color=Country))+
  geom_hline(aes(yintercept= mean(wins)))+
  xlab("Country") +   
  ylab("Number of Gold medalists") +   
  ggtitle("Number of Gold medalists from each country")+
   theme(axis.text.x = element_text(angle=90, hjust=0.5))+
  coord_fixed()
```



### create sd values with plot

We can also use standard deviations as an interpretable unit

```{r sd_cal}
sds_to_plot <- seq(-6,6)
sd_df <- country_count %>%
  summarize(mean_wins = mean(wins), sd_wins = sd(wins)) %>%
  slice(rep_along(sds_to_plot, 1)) %>%
  mutate(sd_to_plot=sds_to_plot) %>%
  mutate(sd_val = mean_wins + sd_to_plot * sd_wins)

country_count %>%
  ggplot(aes(x=wins)) + 
    geom_histogram(bins=30) +
    geom_vline(aes(xintercept=mean(wins)), col="blue", size=2.5) +
    geom_vline(aes(xintercept = sd_val), data=sd_df,
               linetype=2, size=1.2 - abs(seq(-1,1, len=13)))

```

### Sport's popularity based on medals won and it's plot
This line plot gives an outcomeof popular sports or the sports for which mot medals are taken. It looks like aquatics is the popular sport. We will see five orde rsummary to see the skew.

```{r plot_three}
country_lead <- olympics_data%>% 
  select(Country, Sport)%>%
  group_by(Sport)%>%
  summarize(Popular_sport = length(Sport))%>%
  arrange(desc(Popular_sport))
country_lead
```

```{r plot_four}
aquatics<-country_lead%>%
  ggplot(aes(y=Popular_sport, x = Sport, group = 1))+ 
  geom_line(color= "red", size= 1.5)+
  xlab("Sport") +   
  ylab("Number of medals won") +   
  ggtitle("Number of Medals won on particular sport")+
   theme(axis.text.x = element_text(angle=90, hjust=0.5))
aquatics
```

```{r find_skew}
country_lead%>%
  summarize(med_popular = median(Popular_sport), 
            q1_popular = quantile(Popular_sport, 1/4),
            q3_popular = quantile(Popular_sport, 3/4)) %>%
  mutate(d1_depth = med_popular - q1_popular,
         d2_depth = q3_popular - med_popular) %>%
  select(d1_depth, d2_depth)
```

So,from above, we can see that the range of data on one side of the median is longer (or shorter) than the range of data on the other side of the median; therefore, the Popular_sport dataset is skewed.

### Sport's Popularity based on the top five country with highest Medals

We will use facet grid to represent the total number of medals won by top five countries in the Olympics.

```{r plot_2}
country_lead2 <- olympics_data%>% 
  filter(Country %in% c("USA", "GBR","CHN", "RUS", "GER"))%>%
  group_by(Sport, Country)%>%
  summarize(Popular_sport_Count= length(Sport))%>%
  arrange(desc(Popular_sport_Count))%>%
  ggplot(aes(y=Popular_sport_Count, x = Sport))+
  facet_grid(Country~.)+
  geom_point(mapping=aes(color= Country))+
  geom_smooth(method=lm)+
  xlab("Sport") +   
  ylab("Number of medals won") +   
  ggtitle("Number of Medals Won by Top-five countries")+
  theme(axis.text.x = element_text(angle=90, hjust=0.5))
country_lead2
``` 

### Top three players in the whole game 

```{r players_name}
players<-have_record%>%
  select(Athlete_Name, Gender)%>%
  group_by(Athlete_Name, Gender)%>%
  summarize(athlete_count = length(Athlete_Name))%>%
  arrange(desc(athlete_count))
players%>%head(3)
```

### Categories of aquatics winning highest medals

We will plot the graph by reordering it.Until now, the graphs were not sorted but this plot uses reorder function whilch sorts the graph in descending order.

```{r swimming}
swimming <- olympics_data%>%
  filter(Sport == "Aquatics")%>%
  group_by(Event_name)%>%
  summarize(Count = length(Event_name))
  
swimming%>%
  ggplot(aes(x = reorder(Event_name, -Count), y = Count)) + 
  geom_bar(stat = "identity", mapping = aes(fill= Count, color=Count))+   
  scale_x_discrete(labels = function(x) str_wrap(x, width = 5)) +       
  theme(axis.title = element_text(face= "italic", color = "blue", size = 15))+
  labs(x = "Events", y = "Count")+
  ggtitle("Count of medals on different events of Aquatics")+
  coord_flip()

  
```
 

### Analyzing the count of people having world record(WR) and Olympic Record(OR)

```{r wr_and_or}
count_OR <- have_record%>%
  group_by(if_OR)%>%
  summarise(or_count = length(if_OR))

lbls <- ifelse(count_OR$if_OR == "Yes", "Olympic Record", "World Record")
pie(count_OR$or_count , labels = lbls, main = "Pie Chart of Record in olympics")
```

The following data is the subset of have_record datframe and gives the total number of Medals won for world record and olympic record. If the if_OR is Yes, it is olympic record, otherwise, it's world record.

```{r plot_six}
country_record <- have_record%>% 
  filter(Country %in% c("USA", "GBR","CHN", "RUS", "GER"))%>%
  group_by(if_OR, Country)%>%
  subset(select= -c(Sport, Records))%>%
  summarise(Yes_Count = length(if_OR))
country_record
```

### Representation of above data in 3D
This data reveals that among top five countries winning large number of medals, only China, Great Britain and US has people with world record.

```{r piechart}
slices <- c(2, 1, 3)
lbl <- c("China- 2", "Great Britain- 1", "US- 3")
pie3D(slices,labels=lbl,explode=0.1,
  	main="Pie Chart of World Records ")
```

This data reveals that among top five countries winning large number of medals, only China, Great Britain, Germany Russia and US has people with olympic record.

```{r pie_char2}
slices_2 <- c(1, 1, 3, 1, 5)
lbl_2 <- c("China- 1", "Great Britain- 1", "Germany- 3" , "Russia- 1", "US- 5")
pie3D(slices_2,labels=lbl_2,explode=0.1,
  	main="Pie Chart of Olympic Records ")
```

## Hypothesis Testing and Machine learning

We will be performing a statistical  hypothetical testing. Hypothesis testing is, in a way, the formal way of validating the hypothesis claim that is made about a population. Statistical Hypothesis can be classified into into 2 types; null hypothesis and alternative hypothesis.

In simple words, null Hypothesis is the claim that's on trial is called the null hypothesis. The null hypothesis testing is denoted by H0.
Alternative Hypothesis is the one we would believe if the null hypothesis is concluded to be untrue. The evidence in the trial is data and the statistics that go along with it. The alternative hypothesis testing is denoted by H1 or Ha.

The another important term is significance level which basicall means the probability of rejecting the null hypothesis in a statistical test when it is true. Usually 0.05 is taken as the significance level.

### Hypothesis Testing
Through EDA, we are able to find that there is not much Gender differences on winning the medal. The common stereotype is that men are most capable in physical activites and women would lag behind. However, looking at the table above, we don't see much difference in the mean of Male and Female.It is observed that among the top three players, two of them are women. Therefore we can carry out significance tests  using alpha = 0.05. So, the hypothesis for this tutorial is based on the gender differences. The null and alternative hypothesis for this tutorial is stated below:

### Null hypothesis: 
Null hypothesis: There is no difference between males and females in terms of winning medals in olympics .It is symbolized as Ho:??1?????2=0

### Alternative hypothesis: 
Alternative hypotheis: There are differences between males and females in terms of winning medals in olympics. It is symbolized as Ha:??1?????2???0.

### t-test statistics

A t-test looks at the t-statistic, the t-distribution and degrees of freedom to determine the probability of difference between variables.

The basic t- test is applied on this tutorial. We will apply t-test and compare two vectors. We will be analyzing data on the basis of sports involved and the gender.We are using significance level of 5%.

The following data gives count for gender which are using in tidying the dataset. Also, we eliminate the X gender because, they fall on the minority category and we can see that there is huge difference between X gender and other genders.

```{r tree}
# subset of tidy data with Sport, gender, gender count
tree_data <- olympics_data%>%
  select(Sport, Gender)%>%
  group_by(Sport, Gender)%>%
  summarize(g_count = length(Gender))%>%
  subset(Gender!= "X")
tree_data
```

### Descriptive Statistics

Descriptive statitics are often used to describe basic features of data in a study.They are the basis of virtually every quatitave data. Below, we will see some factors taken in account for descriptive statistics and analysis.

```{r tree_stat}
stat_tree <- tree_data%>%
  group_by(Gender)%>%
  summarize(avg_count = mean(g_count), sd_count = sd(g_count))
stat_tree
```

Given dataset below is filtering the tree_data with two datasets having men and women counterparts separately.

```{r tree_divide}
#divide the datset into two Groups
treeM <- tree_data %>%
  filter(Gender == "M")%>%
  group_by(Gender)
  
treeF <- tree_data %>%
  filter(Gender == "W")%>%
  group_by(Gender)
```

we will now perform a t_test with 95% confidence interval and it is going to be two sided. Confidence interval is the range of values defined such that there is specified probability that the value of parameter lies within.

```{r t_test}
t.test(treeM$g_count, treeF$g_count, mu = 0, con4f.level = 0.95, alternative = "two.sided", paired = FALSE)
```

The table shows t = 0.53681, with the degree of freedom df = 51.985. The p-value equals 0.5937. With this given p-value and using significance level of 0.05, we fail to reject the null hypothesis and reject the alternative hypothesis saying that there is a difference between the male and the females. Therefore, we cannot conclude that there is a significant difference between the physical qualities and sport involvement of female and male players.

A 95% confidence interval for the difference in mean count is given by

```{r mean_diff}
diff <- treeM$g_count - treeF$g_count

t.test(diff)
```

Form the above data, the confidence interval is between 0.02 and 1.75. This shows that 95% of the true population mean differences between these two  gender are in this intervals.

## Machine learning

### Linear Regression

One of the most used techniques in class is linear regression where we can examine the potential relationship between variable and a set of oe or more variables.In this tutorial, we will examine linear model for gender count.

Following is the plot which gets a violin plot for the count of gender.

```{r linear_1}
graph_tree<-tree_data %>%
  ggplot(aes(y = g_count, x= Gender)) +
  geom_violin(aes(fill= Gender, color=Gender)) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle=90, hjust=0.5))
  
graph_tree
```
From the above figure,we can observe that the number of men is greater heightwise than women whereas the nuber of women is greater breadthwise than men. So, looking at this figure, it is likely to seem that there is not much difference between the men and women winning in OLympics.

Following code uses lm function to interpret the data.

```{r tree_fit1}
library(broom)
tree_fit <- lm(g_count~Gender, data = tree_data, family = binomial)
tree_fit
```

According to this model, when the female is not counted, the total number of men counterparts would be 5.9630. This statement can be verified from the descriptive statistics too. Also note that the units of outcome of Y and the predictor x matter for the values.

```{r tidy_f}
tidy_tree_fit <- tree_fit%>%
  tidy()
tidy_tree_fit
```
This tidy function is defined in broom package which helps to manipulate the result of learning models in a cothe linearmodel with imformation of the fitted model of observation.

```{r augment}
augmented_gap <- tree_fit%>%
  augment()

augmented_gap%>% head()
```

```{r a_gap}
augmented_gap%>% 
  ggplot(aes(x=factor(Gender), y=.resid)) +
  geom_boxplot() + 
  labs(title="Residuals over Gender", x = "Gender", y = "residual")

```

There seem to be a slight dependence of residuals and Gender as the residuals are less distibuted for women.On the other hand, we can make the plot to check for possible nonlinearity.

```{r a_g}
augmented_gap%>% 
  ggplot(aes(x=.fitted, y=.resid)) +
  geom_point() + 
  geom_smooth(method = loess) +
  labs(title="Residuals over Gender", x = "fitted", y = "residual")

```

We are going to consider anoher linear regresion model which considers Sport and G_count.

```{r two}
tree_fit_two <- lm(g_count~Sport,data=tree_data)%>% 
  augment()
tree_fit_two %>% 
  ggplot(aes(x=factor(Sport), y=.resid)) + 
  geom_boxplot() +
  labs(title="Residuals across Sport over gender", x = "Sport", y = "residuals")+
  theme(axis.text.x = element_text(angle=90, hjust=0.5))
```
The data distributed is unsymmetric and boxing seem tohave high number of residuals.

```{r plot_seven}
tree_data%>%
  ggplot(aes(y= g_count, x = Sport, color = Gender, group = 1))+ 
  geom_point()+ 
  geom_smooth(method = lm) +
  theme(axis.text.x = element_text(angle=90, hjust=0.5))
```

Based on tha above plot, it looks considering the sport would be beneficial to determine the gender involvement in the specified sport.We can observe that women are more involved in many sports than men. But men are ranked higher than wome, Therefore, it is due to women's involvemnt in lot of sports than men make it look that there is not significant disfference between male and female counter parts.

## Summary 

Olympics 2016 draw attention to the fact that women are not less than any boys in physical activities like sports. This has been determined from hypothesis testing and machine learning process. Similarly, Aquatics was the most popular spot and among aquatics, frees tyle event was the highest to get more medals. USA stood in first position with almost double wins than Great Britain. Therefore, this tutorial also proves that a simple dataset can contain lots of information which can be fruitful for future use and provides evidence that  data scraping, EDA and machine learning should be the emerging knowledge preffered in the develoment of data analytics..

## References:

This tutorial highlighted some techniques for data analysis. Much more detail about the libraries and question on IMDB data come be find on the following links.


